Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-11-26 paper/main.tex

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}
\newcommand{\corrmats}{5}
\newcommand{\matchmats}{6}
\newcommand{\kopt}{7}
\newcommand{\arrows}{4}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Supplementary Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Supplementary Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Supp.\ Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Supp.\ Fig.~\corrmats), each participant appeared to have a unique ``recall resolution,'' reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event (Fig.~\ref{fig:model}E, Supp.\ Fig.~\corrmats)~\citep{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Fig.~\ref{fig:model}G, Supp.\ Fig.~\matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first (probability of first recall~\citep{AtkiShif68, PostPhil65, WelcBurn24}; Fig.~\ref{fig:list-learning}A); how participants most often transitioned between recalls of the events as a function of the temporal distance between them (lag-conditional response probability~\citep{Kaha96}; Fig.~\ref{fig:list-learning}B); and which events they were likely to remember overall (serial position recall analyses~\citep{Murd62a}; Fig.~\ref{fig:list-learning}C). Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for an example participant (P17), chosen for their large number of recall events (for analogous figures for other participants, see Supp.\ Fig.~\corrmats).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across the (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The across-participants (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the episode and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original episode (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  For additional detail see \textit{Methods} and Supplementary Figure~\arrows.  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The episode's trajectory is shown in black for reference.  Here, events (dots) are colored by their matched episode event (Panel A).}
\label{fig:trajectory}
\end{figure}

Visual inspection of the episode and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the episode (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in the reduced topic space, could the position of their next recalled event be predicted reliably?  For each location in the reduced topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods}, Supp.\ Fig.~\arrows).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire episode trajectory exhibited such peaks.  In other words, participants' recalls exhibited similar trajectories to each other that also matched the trajectory of the original episode (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of HMM-identified recall events (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the episode.  Differences in the numbers of recall events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of disentangling classic ``proportion recalled'' measures (i.e., the proportion of episode events described in participants' recalls) from participants' abilities to recapitulate the episode's essence (i.e., the similarity between the shapes of the original episode trajectory and that defined by each participant's recounting of the episode).

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Prior studies on narrative structure and processing have shown that we both perceive and internally represent the world around us at multiple, hierarchical timescales~\citep{HassEtal08, LernEtal11, HassEtal15, ChenEtal17, BaldEtal17, BaldEtal18}.  However, for the purposes of our framework, we sought to identify the single timeseries of event representations that was emphasized most heavily in the temporal structure of the episode and of each participant's recall.  We quantified this as the set of $K$ states that maximized the similarity between topic vectors for timepoints comprising each state, while minimizing the similarity between topic vectors for timepoints across different states.  Specifically, we computed (for each matrix)
\[
  \argmax_K \left[W_{1}(a, b)\right],
\]
where $a$ was the distribution of within-state topic vector correlations, and $b$ was the distribution of across-state topic vector correlations .  We computed the first Wasserstein distance ($W_{1}$, also known as ``Earth mover's distance''~\citep{Dobr70, RamdEtal17}) between these distributions for a large range of possible $K$-values (range [2, 50]), and selected the $K$ that yielded the maximum value.  Figure~\ref{fig:model}B displays the event boundaries returned for the episode, and Supplementary Figure~\corrmats~displays the event boundaries returned for each participant's recalls.  See Supplementary Figure \kopt~for the optimization functions for the episode and recalls.  After obtaining these event boundaries, we created stable estimates of the content represented in each event by averaging the topic vectors across timepoints between each pair of event boundaries.  This yielded a number-of-events by number-of-topics matrix for the episode and recalls from each participant.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories; also see Supp.\ Fig.~\arrows).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a $60 \times 60$ (arbitrary units) square.  We tiled this space with a $50 \times 50$ grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B, we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by the line segments that passed within 2.4 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

@@

Next, we constructed a series of ``template'' matrices.  The first template reflected the timecourse of the episode's topic proportions matrix, and the others reflected the timecourse of each participant's recall topic proportions matrix.  To construct the episode template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the topic proportions matrices into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic proportions matrix (Fig.~\ref{fig:model}D, Supp.\ Fig.~\corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic proportions matrices with the episode topic proportions matrix.  An example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B.  This yielded a $1976 \times 1976$ correlation matrix for the episode template and for each participant's recall template.

>>>>

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}

\newcommand{\arrows}{1}
\newcommand{\corrmats}{2}
\newcommand{\matchmats}{3}
\newcommand{\kopt}{4}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Extended Data Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Extended Data Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Extended Data Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Extended Data Fig.~\corrmats), each participant appeared to have a unique ``recall resolution,'' reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event (Fig.~\ref{fig:model}E, Extended Data Fig.~\corrmats)~\citep{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Fig.~\ref{fig:model}G, Extended Data Fig.~\matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first (probability of first recall~\citep{AtkiShif68, PostPhil65, WelcBurn24}; Fig.~\ref{fig:list-learning}A); how participants most often transitioned between recalls of the events as a function of the temporal distance between them (lag-conditional response probability~\citep{Kaha96}; Fig.~\ref{fig:list-learning}B); and which events they were likely to remember overall (serial position recall analyses~\citep{Murd62a}; Fig.~\ref{fig:list-learning}C). Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for an example participant (P17), chosen for their large number of recall events (for analogous figures for other participants, see Extended Data Fig.~\corrmats).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across the (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The across-participants (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the episode and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original episode (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  For additional detail see \textit{Methods} and Extended Data Figure~\arrows.  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The episode's trajectory is shown in black for reference.  Here, events (dots) are colored by their matched episode event (Panel A).}
\label{fig:trajectory}
\end{figure}

Visual inspection of the episode and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the episode (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in the reduced topic space, could the position of their next recalled event be predicted reliably?  For each location in the reduced topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods}, Extended Data Fig.~\arrows).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire episode trajectory exhibited such peaks.  In other words, participants' recalls exhibited similar trajectories to each other that also matched the trajectory of the original episode (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of HMM-identified recall events (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the episode.  Differences in the numbers of recall events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of disentangling classic ``proportion recalled'' measures (i.e., the proportion of episode events described in participants' recalls) from participants' abilities to recapitulate the episode's essence (i.e., the similarity between the shapes of the original episode trajectory and that defined by each participant's recounting of the episode).

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Prior studies on narrative structure and processing have shown that we both perceive and internally represent the world around us at multiple, hierarchical timescales~\citep{HassEtal08, LernEtal11, HassEtal15, ChenEtal17, BaldEtal17, BaldEtal18}.  However, for the purposes of our framework, we sought to identify the single timeseries of event representations that was emphasized most heavily in the temporal structure of the episode and of each participant's recall.  We quantified this as the set of $K$ states that maximized the similarity between topic vectors for timepoints comprising each state, while minimizing the similarity between topic vectors for timepoints across different states.  Specifically, we computed (for each matrix)
\[
  \argmax_K \left[W_{1}(a, b)\right],
\]
where $a$ was the distribution of within-state topic vector correlations, and $b$ was the distribution of across-state topic vector correlations .  We computed the first Wasserstein distance ($W_{1}$, also known as ``Earth mover's distance''~\citep{Dobr70, RamdEtal17}) between these distributions for a large range of possible $K$-values (range [2, 50]), and selected the $K$ that yielded the maximum value.  Figure~\ref{fig:model}B displays the event boundaries returned for the episode, and Extended Data Figure~\corrmats~displays the event boundaries returned for each participant's recalls.  See Extended Data Figure~\kopt~for the optimization functions for the episode and recalls.  After obtaining these event boundaries, we created stable estimates of the content represented in each event by averaging the topic vectors across timepoints between each pair of event boundaries.  This yielded a number-of-events by number-of-topics matrix for the episode and recalls from each participant.

@@

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories; also see Extended Data Fig.~\arrows).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a $60 \times 60$ (arbitrary units) square.  We tiled this space with a $50 \times 50$ grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B, we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by the line segments that passed within 2.4 units.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

@@

Next, we constructed a series of ``template'' matrices.  The first template reflected the timecourse of the episode's topic proportions matrix, and the others reflected the timecourse of each participant's recall topic proportions matrix.  To construct the episode template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the topic proportions matrices into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic proportions matrix (Fig.~\ref{fig:model}D, Extended Data Fig.~\corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic proportions matrices with the episode topic proportions matrix.  An example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B.  This yielded a $1976 \times 1976$ correlation matrix for the episode template and for each participant's recall template.
